Clustering time series based on probability distributions across temporal granularities

### 2021-11-01 main.Rmd
notation of level change

To validate the clustering approaches, we create data designs that replicate prototype behaviors that might be seen in electricity data contexts. We spiked several attributes in the data to see where one method works better than the other and where they might give us the same outcome or the effect of missing data <!---and trends--> on the proposed methods. Three circular granularities $g1$, $g2$ and $g3$ are considered with categories denoted by $\{g10,g11\}$, $\{g20, g21, g22\}$ and $\{g30, g31, g32, g33, g34\}$ and levels $l_{g_1}=2$, $l_{g_2}=3$ and $l_{g_3}=5$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by ${0, 1, 2, \dots, 6}$ or ${Mon, Tue, \dots, Sun}$. Here categories of $g1$, $g2$ and $g3$ are represented by  $\{0, 1\}$, $\{0, 1, 2\}$ and $\{0, 1, 2, 3, 4\}$ respectively. A continuous measured variable $v$ of length $T$ indexed by $\{0, 1, \dots T-1\}$ is simulated such that it follows the structure across $g1$, $g2$ and $g3$. We constructed independent replications of all data designs $R = 25, 250, 500$ to investigate if our proposed clustering method can discover distinct designs in small, medium, and big number of series. All designs employ $T=300, 1000, 5000$ sample sizes to evaluate small, medium, and large sized series. Variations in method performance may be due to different jumps between categories. So a mean difference of $diff = 1, 2, 5$ is examined. The approaches' performance varies with the number of significant granularities. So all, few, and one major granularity scenarios are considered. Please see the Supplementary section (https://github.com/Sayani07/paper-gracsR) for the code and findings.

@@

The proposed methods could be extended when two granularities of interest interact and we want to group subjects based on the interaction of the two granularities. Consider a group that has a different weekday and weekend behavior in the summer but not in the winter. This type of combined behaviour across granularities  can be discovered by evaluating the distribution across combinations of categories for different interacting granularities (wknd-wday and month-of-year in this example). As a result, in this scenario, we analyse a combination of categories generated from different distributions. Consider a case in which there are only two interacting granularities of interest, $g1$ and $g2$. In contrast to the previous situation, when we could study distributions across $l_{g_1} + l_{g_2} = 5$ separate categories, with interaction, we must evaluate the distribution of the $l_{g_1}*l_{g_2}=6$ combination of categories. Consider the $4$ designs in Figure \ref{fig:interaction-gran}, where various distributions are assumed for different combinations of categories, resulting in different designs. Design-1 exhibits no change in distributions across $g1$ or $g2$, whereas Design-2 and Design-3 alter across only $g1$ and $g2$, respectively. Design-4 varies across both $g1$ and $g2$ categories. Design-3 and Design-4 appear similar based on their relative difference across consecutive categories, however Design-4 also changes across facets, unlike Design-3, which has all facets look the same.

>>>>

To validate the clustering approaches, we create data designs that replicate prototype behaviors that might be seen in electricity data contexts. We spiked several attributes in the data to see where one method works better than the other and where they might give us the same outcome or the effect of missing data <!---and trends--> on the proposed methods. Three circular granularities $g1$, $g2$ and $g3$ are considered with categories denoted by $\{g10,g11\}$, $\{g20, g21, g22\}$ and $\{g30, g31, g32, g33, g34\}$ and levels $n_{g_1}=2$, $n_{g_2}=3$ and $l_{g_3}=5$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by ${0, 1, 2, \dots, 6}$ or ${Mon, Tue, \dots, Sun}$. Here categories of $g1$, $g2$ and $g3$ are represented by  $\{0, 1\}$, $\{0, 1, 2\}$ and $\{0, 1, 2, 3, 4\}$ respectively. A continuous measured variable $v$ of length $T$ indexed by $\{0, 1, \dots T-1\}$ is simulated such that it follows the structure across $g1$, $g2$ and $g3$. We constructed independent replications of all data designs $R = 25, 250, 500$ to investigate if our proposed clustering method can discover distinct designs in small, medium, and big number of series. All designs employ $T=300, 1000, 5000$ sample sizes to evaluate small, medium, and large sized series. Variations in method performance may be due to different jumps between categories. So a mean difference of $diff = 1, 2, 5$ is examined. The approaches' performance varies with the number of significant granularities. So all, few, and one major granularity scenarios are considered. Please see the Supplementary section (https://github.com/Sayani07/paper-gracsR) for the code and findings.

@@

The proposed methods could be extended when two granularities of interest interact and we want to group subjects based on the interaction of the two granularities. Consider a group that has a different weekday and weekend behavior in the summer but not in the winter. This type of combined behaviour across granularities  can be discovered by evaluating the distribution across combinations of categories for different interacting granularities (wknd-wday and month-of-year in this example). As a result, in this scenario, we analyse a combination of categories generated from different distributions. Consider a case in which there are only two interacting granularities of interest, $g1$ and $g2$. In contrast to the previous situation, when we could study distributions across $n_{g_1} + n_{g_2} = 5$ separate categories, with interaction, we must evaluate the distribution of the $n_{g_1}*n_{g_2}=6$ combination of categories. Consider the $4$ designs in Figure \ref{fig:interaction-gran}, where various distributions are assumed for different combinations of categories, resulting in different designs. Design-1 exhibits no change in distributions across $g1$ or $g2$, whereas Design-2 and Design-3 alter across only $g1$ and $g2$, respectively. Design-4 varies across both $g1$ and $g2$ categories. Design-3 and Design-4 appear similar based on their relative difference across consecutive categories, however Design-4 also changes across facets, unlike Design-3, which has all facets look the same.
